Predator clusters

### 2023-03-22 reports/04_pred_clus.qmd
NMDS and ice categories

library(ggmosaic)
library(ggtree)
library(sf)
library(targets)
library(tidyverse)
library(vegan)

tar_load(c("predators_stations", "predators_abundant", "sightings_clust",
           "sightings_mtx", "sightings_dist", "zoop_sf"), 
         store = here::here("_targets"))
source(here::here("R", "sp.R"))
source(here::here("R", "zoop.R"))
```

@@

We use the gap statistic [@tibshirani2001] to choose the optimal number of clusters ($k$). The gap statistic ($f$) is a goodness of clustering measure. The authors recommended choosing the smallest $k$ such that $f(k) \geq f(k+1) - s(k+1)$ where $s$ is the standard error of $f$. According to this rule, the predators are best described by two clusters. However, $f(2)$ is only slightly greater than $f(3)-s(3)$ and, because $s$ is estimated by bootstrapping, $k=3$ may be selected sometimes depending on stochasticity.

```{r}
cluster_fn <- function(x, k) {
  bray_dist <- function(x) {
    vegdist(x, method = "bray")
  }
  list(cluster = cutree(hclust(bray_dist(x), method = "ward.D2"), k = k))
}
set.seed(931)
sightings_gap <- suppressWarnings(
  cluster::clusGap(sightings_mtx, 
                   FUN = cluster_fn,
                   K.max = 10, 
                   B = 500,
                   verbose = FALSE)
)
sightings_gap_df <- as_tibble(sightings_gap$Tab) %>% 
  mutate(
    k = seq_along(logW),
    is_optimal = k == cluster::maxSE(gap, SE.sim, method = "Tibs2001SEmax")
  )
ggplot(sightings_gap_df, aes(x = k, y = gap)) +
  geom_line() +
  geom_pointrange(aes(ymin = gap - SE.sim, ymax = gap + SE.sim,
                      color = is_optimal)) +
  scale_color_manual(values = c(`TRUE` = "red",
                                `FALSE` = "black")) +
  theme_classic() +
  theme(legend.position = "none")
```

## 2-cluster analysis

Repeat the indicator species analysis and maps with two clusters.

@@

# Cut predators into 2 clusters
clust <- cutree(sightings_clust, 2)
g <- split(names(clust), clust)
p <- ggtree(sightings_clust, hang = -1)
pred_clust_mrca <- sapply(g, function(n) MRCA(p, n))

@@

@@

```{r}
#| label: nmds

sightings_nmds <- metaMDS(
  sightings_mtx, 
  distance = "bray",
  k = 3,
  trymax = 100
)

stressplot(sightings_nmds)

@@

@@

ggplot(nmds_df, aes(NMDS1, NMDS2)) +
  geom_point(aes(color = pred_clust)) +
  theme_classic()

ggplot(nmds_df, aes(NMDS1, NMDS3)) +
  geom_point(aes(color = pred_clust)) +
  theme_classic()

ggplot(nmds_df, aes(NMDS2, NMDS3)) +
  geom_point(aes(color = pred_clust)) +
  theme_classic()
```

>>>>

library(cluster)
library(ggmosaic)
library(ggtree)
library(sf)
library(targets)
library(tidyverse)
library(vegan)

tar_load(c("predators", "predators_stations", 
           "predators_abundant", "sightings_clust", "sightings_mtx", 
           "sightings_dist", "zoop_sf"), 
         store = here::here("_targets"))
source(here::here("R", "sp.R"))
source(here::here("R", "zoop.R"))
```

@@

We use the gap statistic [@tibshirani2001] to choose the optimal number of clusters ($k$). The gap statistic ($f$) is a goodness of clustering measure. The authors recommended choosing the smallest $k$ such that $f(k) \geq f(k+1) - s(k+1)$ where $s$ is the standard error of $f$. According to this rule, the predators are best described by two clusters (indicated in blue). However, $f(2)$ is only slightly greater than $f(3)-s(3)$ and, because $s$ is estimated by bootstrapping, sometimes re-running the gap statistic estimate yields an optimum at $k=3$ instead. An alternative rule suggests choosing the smallest $k$ such $f(k) \geq f_{max} - s_{max}$, where $f_{max}$ and $s_{max}$ are $f$ and $s$ at the first local maximum, respectively. In the figure below, $f_{max} - s_{max}$ is represented by the dashed red line, and the smallest $k$ greater than that threshold is $k=3$ (red).

```{r}
cluster_fn <- function(x, k) {
  bray_dist <- function(x) {
    vegdist(x, method = "bray")
  }
  list(cluster = cutree(hclust(bray_dist(x), method = "ward.D2"), k = k))
}
set.seed(931)
sightings_gap <- suppressWarnings(
  clusGap(sightings_mtx, 
                   FUN = cluster_fn,
                   K.max = 10, 
                   B = 500,
                   verbose = FALSE)
)
sightings_gap_df <- as_tibble(sightings_gap$Tab) %>% 
  mutate(
    k = seq_along(logW),
    is_optimum = case_when(
      k == maxSE(gap, SE.sim, method = "Tibs2001SEmax") ~ "tibsSEmax",
      k == maxSE(gap, SE.sim, method = "firstSEmax") ~ "firstSEmax",
      TRUE ~ "none"
  ))
kfirstmax <- with(
  sightings_gap_df,
  maxSE(gap, SE.sim, method = "firstmax")
)
ktibs <- with(sightings_gap_df, k[is_optimum == "tibsSEmax"])
firstSEmax_thr <- with(sightings_gap_df, gap[kfirstmax] - SE.sim[kfirstmax])
tibs_thr <- with(sightings_gap_df, gap[ktibs + 1] - SE.sim[ktibs + 1])
ggplot(sightings_gap_df, aes(x = k, y = gap)) +
  geom_line() +
  geom_pointrange(aes(ymin = gap - SE.sim, ymax = gap + SE.sim,
                      color = is_optimum)) +
  geom_segment(x = 3, xend = 4, 
               y = firstSEmax_thr, yend = firstSEmax_thr,
               color = "red",
               linetype = "dashed") +
  geom_segment(x = ktibs, xend = ktibs + 1, 
               y = tibs_thr, yend = tibs_thr,
               color = "blue",
               linetype = "dashed") +
  scale_color_manual(values = c(tibsSEmax = "blue",
                                firstSEmax = "red",
                                none = "black")) +
  theme_classic() +
  theme(legend.position = "none")
```

## 3-cluster analysis

Repeat the indicator species analysis and maps with three clusters.

@@

# Cut predators into 3 clusters
clust <- cutree(sightings_clust, 3)
g <- split(names(clust), clust)
p <- ggtree(sightings_clust, hang = -1)
pred_clust_mrca <- sapply(g, function(n) MRCA(p, n))

@@

ant_basemap(map_lim) +
  geom_sf(aes(color = pred_clust), pred_clust_sf, size = 2, shape = 15) +
  facet_wrap(~ Year) +
  scale_x_continuous(breaks = c(-60, -55)) +
  scale_y_continuous(breaks = c(-63, -62, -61, -60)) +
  scale_color_brewer(palette = "Dark2") +
  coord_ant(map_lim) +
  theme(legend.position = "bottom")
```

@@

How many NMDS axes should we use? Scree plot

```{r}
#| label: nmds-scree

nmds_stress <- tibble(
  k = 1:6,
  stress = map_dbl(k, function(k) {
    metaMDS(
      sightings_mtx, 
      distance = "bray",
      k = k,
      trymax = 100,
      trace = FALSE
    )$stress
  })
)

ggplot(nmds_stress, aes(k, stress)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = c(0.2, 0.1, 0.05), linetype = "dotted") +
  scale_x_continuous(breaks = nmds_stress$k) +
  expand_limits(y = 0) +
  theme_classic()

@@

3 axes for NMDS looks acceptable. How does ordination distance compare to dissimilarity (Shepard diagram)?

```{r}
#| label: nmds-shepard

sightings_nmds <- metaMDS(
  sightings_mtx, 
  distance = "bray",
  k = 3,
  trymax = 100,
  trace = FALSE
)

nmds_isoreg <- isoreg(sightings_nmds$diss, sightings_nmds$dist)

nmds_shepard <- tibble(
  diss = sightings_nmds$diss,
  dist = sightings_nmds$dist,
  mono = nmds_isoreg$yf
)

nmds_spear <- cor(nmds_shepard$diss,
                  nmds_shepard$dist,
                  method = "spearman")
nmds_label <- str_glue(
  "Non-metric fit, R^2 = {round(1 - sightings_nmds$stress^2, 3)}\n",
  "Linear fit, R^2 = {round(nmds_spear^2, 3)}\n",
  "k = {sightings_nmds$ndim}\n",
  "Stress = {round(sightings_nmds$stress, 3)}"
)

ggplot(nmds_shepard, aes(diss, dist)) +
  geom_point(size = 0.75, alpha = 0.1) +
  geom_line(aes(y = mono), color = "red") +
  annotate("text", x = 0.05, y = 2.9, label = nmds_label,
           hjust = 0, vjust = 1) +
  expand_limits(x = 0, y = 0) +
  labs(x = "Observed dissimilarity",
       y = "Ordination distance") +
  theme_classic()

NMDS plots, color-coded by predator cluster.

@@

ggplot(nmds_df, aes(NMDS1, NMDS2)) +
  geom_point(aes(color = pred_clust)) +
  scale_color_brewer(palette = "Dark2") +
  theme_classic()

ggplot(nmds_df, aes(NMDS1, NMDS3)) +
  geom_point(aes(color = pred_clust)) +
  scale_color_brewer(palette = "Dark2") +
  theme_classic()

ggplot(nmds_df, aes(NMDS2, NMDS3)) +
  geom_point(aes(color = pred_clust)) +
  scale_color_brewer(palette = "Dark2") +
  theme_classic()
```

## \[Insert Frozen or Vanilla Ice pun here\]

Get the modal ice type (multi-year, first-year, thin, or open) for each station from the along-track observations. Seems weird that no stations shook out as multi-year ice, right? This figure shows how frequent the modal ice type was at each station. The two grey lines represent 1:1 and 2:1. Stations falling on the upper line had *uniform* ice category observations i.e., every interval within 15 km had the same ice category. Stations falling below the lower line had *highly mixed* ice category observations i.e., the most frequent ice category was observed at less than half the intervals within 15 km of the station. Most points on the figure include multiple stations, indicated by the color. For example, in the THIN ice category (bottom), 8 stations (red point) had 4 intervals with ice category observations where all 4 were THIN (x,y = 4,4).

```{r}
#| label: ice

# Most frequent value i.e., mode
# Note: returns a character
mfv <- function(x) {
  x_tbl <- table(x)
  names(x_tbl)[which.max(x_tbl)]
}
underway_ice <- predators_stations %>% 
  left_join(transmute(predators,
                      year = as.numeric(substr(cruise, 5, 8)),
                      interval,
                      ice_type_desc),
            by = c("year", "interval"),
            multiple = "all") %>% 
  filter(ice_type_desc != "NULL") %>% 
  left_join(read_csv(here::here("data", "ICECAT.csv"),
                     show_col_types = FALSE),
            by = "ice_type_desc") %>% 
  # Ice observations were unique to intervals, but intervals are
  # repeated if there were multiple predator sightings
  group_by(amlr.station,
           interval) %>% 
  summarize(interval_ice_cat = first(ice_cat),
            .groups = "drop_last") %>% 
  summarize(station_ice_cat = mfv(interval_ice_cat),
            n_obs = n(),
            n_mode = sum(interval_ice_cat == station_ice_cat))

underway_ice %>% 
  group_by(n_obs, n_mode, station_ice_cat) %>% 
  summarize(Stations = n(), .groups = "drop") %>% 
  ggplot(aes(n_obs, n_mode, color = Stations)) +
  geom_abline(slope = c(1, 0.5), intercept = 0, color = "grey50") +
  geom_point() +
  labs(x = "Count of intervals with ice category observation",
       y = "Count of intervals with most frequent ice category") +
  scale_color_fermenter(palette = "RdYlBu",
                        breaks = c(1, 2, 3, 5, 8)) +
  coord_fixed() +
  facet_wrap(~ station_ice_cat, ncol = 2) +
  theme_classic()
